Layer dependence as a measure of local dependence

### 2015-02-17 ld7.tex
final changes

\begin{abstract}
A new measure of local dependence called  ``layer dependence" is analysed and illustrated.  Layer dependence measures dependence between two variables at different percentiles in their joint distributions. Layer dependence satisfies coherence properties similar to Spearman's correlation, such as lying between $-1$ and $1$, with $-1$, $0$ and $1$ corresponding to countermonotonicity, independence and comonotonicity, respectively. Spearman's correlation is  a weighted average of layer dependence at different percentiles.  Alternate overall correlation measures are arrived by varying the weights.  Layer dependence allows copulas to be  fitted and tailored to data and expert opinion on the dependence structure.
\end{abstract}

@@

Remaining sections are structured as follows. Section \aref{sintroduction} defines and analyzes layer dependence. Section \aref{sldcurve} demonstrates the appropriateness of layer dependence as a local dependence measure using several copulas. Section \aref{sdecompose} explains the behaviour of layer dependence by decomposing it into a negative function of discordance and dispersion. Section \aref{scoherence} describes coherence properties of layer dependence. Links to existing literature are highlighted in  \sref{sliterature}. Further properties of layer dependence are highlighted in  \sref{sproperties}. Section \aref{smodeling} suggests approaches to simulate a copula given the layer dependence curve, and  \sref{sfitting} applies layer dependence to copula modeling, using historical stock returns as an illustration. Section \aref{saltoverall} discuss alternative coherent measures of overall dependence based on weighted averages of layer dependence. Section \aref{sconclusion} concludes.

@@

The $\alpha$--layer dependence  between $v$ and the $\alpha$--layer of $u$ is
\begin{equation}\label{definition}
\ell_\alpha \equiv \frac{\cov\{v,(u>\alpha)\}}{\cov\{u,(u>\alpha)\}}
=\frac{\cor\{v,(u>\alpha)\}}{\cor\{u,(u>\alpha)\}} \cq 0\leq\alpha\leq 1\;,
\end{equation}
where cov and cor calculate covariance and correlation, respectively, and $(\alpha\le u)$ is the indicator function of the event in brackets. The denominators in \eref{definition} are independent of $C$ and implies $\ell_\alpha=1$ if $u=v$ and $\ell_\alpha=-1$ if $u=1-v$.  Further $-1\leq \ell_\alpha \leq 1$. Independence implies $\ell_\alpha=0$. Coherence properties of $\ell_\alpha$ are explored in \sref{scoherence}.

@@

Layer dependence relates  to the identity
\begin{equation}\label{decompose}
u= \int_0^1 ( u > \alpha )\de\alpha\ .
\end{equation}
 To prove \eref{decompose} note the left hand side is $\int_0^u 1\de \alpha$ equaling the right hand side.  Informally, the derivative of the  ``$\alpha$--layer'' of $u$,  $(\alpha\le u) \de \alpha$ with respect to $u$ is $\infty\de\alpha=1$ when $u=\alpha$ and $0$ otherwise. Thus the $\alpha$--layer captures movements in $u$ solely at $\alpha$, and ignores movements elsewhere. Thus \eref{decompose} shows $u$ is formed from infinitely many $\alpha$--layers, with each layer capturing the variability of $u$ at a different $\alpha$.  Taking expectations of \eref{decompose} yields the familiar result expressing expectation as the integral of the survival function.

Using  \eref{definition} and \eref{decompose}, Spearman's correlation between $u$ and $v$ is
\begin{equation}\label{wtdaverage}
\rho_S\equiv \cor(u,v) =\frac{\cov(u,v)}{\var(u)}=\frac{\int _0^1\cov\{v,(u>\alpha)\}\de\alpha}{1/12}
=\Ex(\ell_\alpha)\; .
\end{equation}
where the expectation $\Ex$ is calculated over $0\le\alpha\le 1$ using density $6\alpha(1-\alpha)$.  The density integrates to $1$, has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Modifying the density leads to different emphasis on different areas of the relationship between $u$ and $v$, and yields alternative measures of overall dependence. For example using the density $n\alpha^{n-1}$ where $n>0$ puts increasing weight on upper tail dependence. This is further explored in \sref{saltoverall}.

@@

The nine panels in \fref{fillustration} display $(u,v)$ scatterplots of exchangeable copulas, and their layer dependence curves $\ell_\alpha$. Each copula has Spearman's correlation $\rho_S=0.6$.

@@

With Archimedean copulas \citep{mcneil2005qrm} $C(\alpha,\beta) =\psi^-\left\{\psi(\alpha)+\psi(\beta)\right\}$ where $\psi$ is the generator function and $\psi^-$ its inverse. In this case closed form expressions for the integrals and hence $\ell_\alpha$ do not exist.

@@

\section{Alternative measures of overall dependence}\label{saltoverall}

@@

The density $w_\alpha$ specifies ``importance" of dependence at percentile $\alpha$. With Spearman's correlation, $w_\alpha=6\alpha(1-\alpha)$: importance is symmetric and highest at the median, and decreases to zero at both tails. Hence $\rho_S$ emphasizes dependence at moderate percentiles and diminishes dependence in the tails. This weighting may be inappropriate when tail dependence is prominent and carries adverse consequences, such as multiple market crashes or catastrophe insurance losses. Alternative densities are described below.

The following densities $w_\alpha$ yield alternative rank dependence measures:
\begin{itemize}

@@

Layer dependence  captures dependence structures in bivariate copulas, and satisfies coherence properties. Taking weighted averages of layer dependence curves yields Spearman's correlation and alternative overall dependence measures.

>>>>

\begin{abstract}
A new measure of local dependence called  ``layer dependence" is analysed and illustrated.  Layer dependence measures dependence between two variables at different percentiles in their joint distribution. Layer dependence satisfies coherence properties similar to Spearman's correlation, such as lying between $-1$ and $1$, with $-1$, $0$ and $1$ corresponding to countermonotonicity, independence and comonotonicity, respectively. Spearman's correlation is  a weighted average of layer dependence at different percentiles.  Alternate overall correlation measures are arrived by varying the weights.  Layer dependence allows copulas to be  fitted and tailored to data and expert opinion on the dependence structure.
\end{abstract}

@@

Remaining sections are structured as follows. Section \aref{sintroduction} defines and analyzes layer dependence. Section \aref{sldcurve} demonstrates the appropriateness of layer dependence as a local dependence measure using several copulas. Section \aref{sdecompose} explains the behaviour of layer dependence by decomposing it into a negative function of discordance and dispersion. Section \aref{scoherence} describes coherence properties of layer dependence. Links to existing literature are highlighted in  \sref{sliterature}. Further properties of layer dependence are described in  \sref{sproperties}. Section \aref{sfitting} applies layer dependence to copula modeling, and uses historical stock returns as an illustration. Section \aref{saltoverall} discuss alternative coherent measures of overall dependence based on weighted averages of layer dependence. Section \aref{sconclusion} concludes.

@@

The $\alpha$--layer dependence  between $v$ and the $\alpha$--layer of $u$ is
\begin{equation}\label{definition}
\ell_\alpha \equiv \frac{\cov\{v,(u>\alpha)\}}{\cov\{u,(u>\alpha)\}}
=\frac{\cor\{v,(u>\alpha)\}}{\cor\{u,(u>\alpha)\}} \cq 0\leq\alpha\leq 1\;,
\end{equation}
where cov and cor calculate covariance and correlation, respectively, and $(u>\alpha)$ is the indicator function of the event in brackets. The denominators in \eref{definition} are independent of $C$ and implies $\ell_\alpha=1$ if $u=v$ and $\ell_\alpha=-1$ if $u=1-v$.  Further $-1\leq \ell_\alpha \leq 1$. Independence implies $\ell_\alpha=0$. Coherence properties of $\ell_\alpha$ are explored in \sref{scoherence}. The $\alpha$--layer of $u$ is further discussed below.

@@

The definition of layer dependence in \eref{definition} exploits the identity
\begin{equation}\label{decompose}
u= \int_0^1 ( u > \alpha )\de\alpha\ .
\end{equation}
 To prove \eref{decompose}, note the left hand side is $\int_0^u 1\de \alpha$ equaling the right hand side. In addition taking expectations of \eref{decompose} yields the familiar result expressing expectation as the integral of the survival function. Informally, the derivative of the  ``$\alpha$--layer'' of $u$,  $(u>\alpha) \de \alpha$ with respect to $u$ is $\infty\de\alpha=1$ when $u=\alpha$ and $0$ otherwise. Thus the $\alpha$--layer captures movements in $u$ solely at $\alpha$, and ignores movements elsewhere. In addition \eref{decompose} shows $u$ is formed from infinitely many $\alpha$--layers, with each layer capturing the variability of $u$ at a different $\alpha$. Calculating covariance between the $\alpha$--layer of $u$ and $v$ and scaling by the same where $u=v$ yields layer dependence.

Using  \eref{definition} and \eref{decompose}, Spearman's correlation between $u$ and $v$ is
\begin{equation}\label{wtdaverage}
\rho_S\equiv \cor(u,v) =\frac{\cov(u,v)}{\var(u)}=\frac{\int _0^1\cov\{v,(u>\alpha)\}\de\alpha}{1/12}
=\Ex(\ell_\alpha)\; .
\end{equation}
where the expectation $\Ex$ is calculated over $0\le\alpha\le 1$ using density $6\alpha(1-\alpha)$.  The density integrates to $1$, has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Modifying the density leads to different emphasis on different areas of the relationship between $u$ and $v$, and yields alternate measures of overall dependence. For example using the density $n\alpha^{n-1}$ where $n>0$ puts increasing weight on upper tail dependence. This is further explored in \sref{saltoverall}.

@@

The nine panels in \fref{fillustration} display $(u,v)$ scatterplots of exchangeable copulas, and their layer dependence curves $\ell_\alpha$ for all $0\le\alpha\le 1$. Each copula has Spearman's correlation $\rho_S=0.6$.

@@

With Archimedean copulas \citep{mcneil2005qrm} $C(\alpha,\beta) =\psi^-\left\{\psi(\alpha)+\psi(\beta)\right\}$ where $\psi$ is the generator function and $\psi^-$ its inverse. In this case closed form expressions for the integrals and hence for $\ell_\alpha$ do not exist.

@@

\section{Alternate measures of overall dependence}\label{saltoverall}

@@

The density $w_\alpha$ specifies ``importance" of dependence at percentile $\alpha$. With Spearman's correlation, $w_\alpha=6\alpha(1-\alpha)$: importance is symmetric and highest at the median, and decreases to zero at both tails. Hence $\rho_S$ emphasizes dependence at moderate percentiles and diminishes dependence in the tails. This weighting may be inappropriate when tail dependence is prominent and carries adverse consequences, such as multiple market crashes or catastrophe insurance losses. Alternate densities are described below.

The following densities $w_\alpha$ yield alternate rank dependence measures:
\begin{itemize}

@@

Layer dependence  captures dependence structures in bivariate copulas, and satisfies coherence properties. Taking weighted averages of layer dependence curves yields Spearman's correlation and alternate overall dependence measures.
